---
title: "Multi-scale Population Estimates"
output:
  html_notebook:
    toc: yes
    toc_float: yes
  html_document:
    df_print: paged
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(fs)
library(glue)
library(readr)
library(dplyr)
library(tidyr)
library(knitr)
library(purrr)
library(sf)
library(ggplot2)
library(scales)
```

Create and define a folder for data 

```{r}
data_dir <- "census-pop-data"
if(!dir.exists(data_dir)){fs::dir_create(data_dir)}
```

# Sub-county Population Data

Download 2000-2010 data

```{r}
citytown_filepath <- "https://www2.census.gov/programs-surveys/popest/datasets/2000-2010/intercensal/cities/sub-est00int.csv"
citytown_file_local <- glue::glue("{data_dir}/{basename(citytown_filepath)}")
if(!file.exists(citytown_file_local)){
  download.file(citytown_filepath, destfile = citytown_file_local)
}
```

Download 2010-2018 data

```{r}
citytown_filepath2 <- "https://www2.census.gov/programs-surveys/popest/datasets/2010-2018/cities/totals/sub-est2018_all.csv"
citytown_file2_local <- glue::glue("{data_dir}/{basename(citytown_filepath2)}")
if(!file.exists(citytown_file2_local)){
  download.file(citytown_filepath2, destfile = citytown_file2_local)
  }
```

Make a function to read in and reshape 

```{r}
reshape_pop_data <- function(citytown_file_local){
  citytown_long <- readr::read_csv(citytown_file_local) %>%
    dplyr::group_by(SUMLEV, STATE, COUNTY, PLACE, COUSUB, NAME, STNAME) %>%
    # convert columns to rows
    tidyr::pivot_longer(cols = dplyr::contains(match = c("POPESTIMATE")), 
                      names_to = "estimate", values_to = "population") %>%
    dplyr::select(SUMLEV, STATE, COUNTY, PLACE, COUSUB, NAME, STNAME, estimate, population) %>%
  # separate year and estimate type into separate columns 
  dplyr::mutate(year = stringr::str_extract(estimate, "\\d+")) %>%
  dplyr::mutate(year = as.numeric(year)) %>%
  dplyr::select(-estimate)
  return(citytown_long)
}

```

Download population data for metropolitan and micropolitan areas

```{r}
cbsa_url <- "https://www2.census.gov/programs-surveys/popest/datasets/2010-2018/metro/totals/cbsa-est2018-alldata.csv"
cbsa_file_local <- glue::glue("{data_dir}/{basename(cbsa_url)}")
if(!file.exists(cbsa_file_local)){
  download.file(cbsa_url, destfile = cbsa_file_local)
  }
```
```{r}
# read in and reshape to fit with subcounty data
cbsa_pop <- cbsa_file_local %>%
  readr::read_csv() %>%
  dplyr::select(1:16) %>%
  tidyr::pivot_longer(cols = dplyr::contains(match = c("POPESTIMATE")), 
                      names_to = "estimate", values_to = "population") %>%
  dplyr::mutate(year = stringr::str_extract(estimate, "\\d+")) %>%
  dplyr::mutate(year = as.numeric(year), SUMLEV = "320") %>%
  dplyr::mutate(STATE = NA, COUNTY = NA, PLACE = NA, COUSUB = NA, STNAME = NA) %>%
  dplyr::select(SUMLEV, STATE, COUNTY, PLACE, COUSUB, NAME, population, year)
```


Make one data frame with 3 datasets combined (CBSA, 2000-2010 subcounty, 2010-2018 subcounty)

```{r}
popdata_long <- list(citytown_file_local, citytown_file2_local) %>%
  purrr::map(~reshape_pop_data(.x)) %>% 
  bind_rows() %>% 
  bind_rows(cbsa_pop)
```

```{r, include=FALSE, echo=FALSE, results="asis"}
head(popdata_long) %>% knitr::kable()
```

# Download spatial data

```{r}
# Counties
county_shp <- glue("{data_dir}/tl_2019_us_county.shp")
if(!file.exists(county_shp)){
  county2019_url <- 'https://www2.census.gov/geo/tiger/TIGER2019/COUNTY/tl_2019_us_county.zip' 
  tmp <- tempfile()
  download.file(county2019_url, destfile = tmp)
  unzip(zipfile = tmp, exdir = data_dir)
}
# Core Based Statistical Areas
cbsa_shp <- glue("{data_dir}/tl_2019_us_cbsa.shp")
if(!file.exists(cbsa_shp)){
  cbsa2019_url <- 'https://www2.census.gov/geo/tiger/TIGER2019/CBSA/tl_2019_us_cbsa.zip' 
  tmp <- tempfile()
  download.file(cbsa2019_url, destfile = tmp)
  unzip(zipfile = tmp, exdir = data_dir)
}
# Places in MD (State code 24)
md_places_shp <- glue("{data_dir}/tl_2019_24_place.shp")
if(!file.exists(md_places_shp)){
  md_places_url <- 'https://www2.census.gov/geo/tiger/TIGER2019/PLACE/tl_2019_24_place.zip'
  tmp <- tempfile()
  download.file(md_places_url, destfile = tmp)
  unzip(zipfile = tmp, exdir = data_dir)
}
# Water Areas for MD (State code 24)
if(!any(fs::dir_ls(data_dir, recurse = TRUE, glob = "areawater"))){
  system(glue('wget -r -np -A "tl_2019_24*_areawater.zip" ftp://ftp2.census.gov/geo/tiger/TIGER2019/AREAWATER -P {data_dir}'))
  zipfiles <- fs::dir_ls(data_dir, recurse = TRUE, glob = "*areawater.zip")
  purrr::walk(zipfiles, ~unzip(zipfile = .x, exdir = data_dir))
}

```

# Where is SESYNC?

```{r, message=FALSE}
counties_sf <- sf::st_read(county_shp)
counties_md <- dplyr::filter(counties_sf, STATEFP == "24")
cbsa_sf <- sf::st_read(cbsa_shp)
places_sf <- sf::st_read(md_places_shp)

sesync <- st_sfc(st_point(
  c(-76.503394, 38.976546)),
  crs = st_crs(counties_sf))

sesync_county_sf <- counties_sf[sesync,]
sesync_cbsa_sf <- cbsa_sf[sesync,]
sesync_place_sf <- places_sf[sesync,]

sesync_county <- sesync_county_sf[["NAMELSAD"]] %>% as.character()
sesync_place <- sesync_place_sf[["NAMELSAD"]] %>% as.character()
sesync_cbsa <- sesync_cbsa_sf[["NAME"]] %>% as.character()

sesync_ids <- data.frame(SUMLEV = c("050", "157", "320"),
                         NAME = c(sesync_county, sesync_place, sesync_cbsa),
                         stringsAsFactors = FALSE)
```

```{r, include=FALSE, echo=FALSE, results="asis"}
knitr::kable(sesync_ids)
```

Use this info to filter the population data

```{r}
# The key for SUMLEV is as follows:
# 040 = State
# 050 = County
# 061 = Minor Civil Division
# 071 = Minor Civil Division place part
# 157 = County place part
# 162 = Incorporated place
# 170 = Consolidated city
# 172 = Consolidated city -- place within consolidated city
# 320 = Core based statistical areas

sesync_ids <- data.frame(SUMLEV = c("050", "157", "320"),
                         NAME = c(as.character(sesync_county_sf[["NAMELSAD"]]),
                                  as.character(sesync_place_sf[["NAMELSAD"]]),
                                  as.character(sesync_cbsa_sf[["NAME"]])))
sesync_pops <- popdata_long %>% 
    dplyr::filter(SUMLEV %in% sesync_ids[["SUMLEV"]]) %>%
    ungroup() %>% 
    dplyr::group_split(SUMLEV) %>%
    purrr::map(~dplyr::inner_join(.x, sesync_ids)) %>% 
  bind_rows() %>% 
  # get rid of Annapolis city, Missouri!
  dplyr::filter(STATE == "24" | is.na(STATE)) 
```

# Plots 

```{r}
# pop_trends <-
  sesync_pops %>%
  ggplot(aes(x = year, y = population, group = NAME)) +
  geom_line(aes(col = NAME), lwd = 2) +
  facet_wrap(vars(NAME), scales = "free_y", ncol = 1) +
  theme_dark() +
  theme(legend.position = "none") +
  xlab(element_blank()) +
  ylab(element_blank()) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_manual(values = c(
    `Baltimore-Columbia-Towson, MD` = "purple",
    `Anne Arundel County` = "red",
    `Annapolis city` = "black"
  )) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Population Estimates")

# ggsave("popdata.png", 
#        pop_trends, 
#        bg = "transparent", 
#        width = 4.5, height = 5, units = "in")
```

# Maps

```{r}
# make list of water area polygons intersecting Maryland
water_sfs_list <- list.files(recursive = TRUE, full.names = TRUE, pattern = "_areawater.shp$") %>%
  purrr::map(~st_read(.x, quiet = TRUE)) %>% 
  purrr::map(~st_union(.x)) %>%
  purrr::map(~st_intersection(.x, counties_md))
```

```{r}
md_map <-
  sesync_cbsa_sf %>%
  ggplot() +
  geom_sf(data = counties_md, fill = "white") +
  geom_sf(fill = "purple", alpha = 0.5, col = "purple", lwd = 2) +
  geom_sf(data = sesync, pch = 4, col = "green") +
  theme_minimal()  +
  theme(axis.text = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA))

for (i in 1:length(water_sfs_list)) { 
  md_map <- md_map + geom_sf(data = water_sfs_list[[i]], fill = "dodger blue", lwd = 0)
}

md_map <- md_map + geom_sf(data = sesync_cbsa_sf, fill = NA, col = "purple", lwd = 2)
md_map

# ggsave("md.png", md_map, bg = "transparent")

```

```{r}
cbsa_map <-
  sesync_county_sf %>%
  ggplot() +
  geom_sf(data = counties_md, fill = "white") +
  geom_sf(fill = "red", col = "red", lwd = 2, alpha = 0.5) +
  geom_sf(data = sesync, pch = 4, col = "green") +
  theme_minimal()  +
  theme(axis.text = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA))

for (i in 1:length(water_sfs_list)) { 
  cbsa_map <- cbsa_map + geom_sf(data = water_sfs_list[[i]], fill = "dodger blue", lwd = 0)
}

cbsa_map <- cbsa_map + geom_sf(data = sesync_county_sf, fill = NA, col = "red", lwd = 2)
cbsa_map
# ggsave("cbsa_map..png", cbsa_map, bg = "transparent")
```

```{r, message=FALSE}
county_bbox <- sesync_county_sf %>% st_bbox()

map1 <- sesync_county_sf %>%
  ggplot() +
  geom_sf(fill = "white", col = "red") +
  geom_sf(data = sesync_place_sf, fill = "black", col = "black", alpha = 0.5) +
  geom_sf(data = sesync, pch = 4, col = "green") +
  theme_minimal()  +
  theme(axis.text = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())

for (i in 1:length(water_sfs_list)) { 
  map1 <- map1 + geom_sf(data = water_sfs_list[[i]], fill = "dodger blue", lwd = 0)
}

# aa_county <- 
map1 +  
  geom_sf(data = sesync_place_sf, fill = NA, col = "black", lwd = 1.5) +
  coord_sf(xlim = c(county_bbox[1], county_bbox[3]),
           ylim = c(county_bbox[2], county_bbox[4])) +
  theme(panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "transparent",colour = NA),
    plot.background = element_rect(fill = "transparent",colour = NA))

# ggsave("aacounty.png", aa_county, bg = "transparent")
```


# Interactive map

```{r}
library(leaflet)

tiger_wms_url <- "https://tigerweb.geo.census.gov/arcgis/services/TIGERweb/tigerWMS_Current/MapServer/WMSServer"

leaflet() %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  setView(lng = -76.5, lat = 38.97, zoom = 8) %>%
  addWMSTiles(
    tiger_wms_url,
    layers = c("Metropolitan Statistical Areas",
               "Metropolitan Statistical Areas Labels",
               "Counties", "Counties Labels",
               "Incorporated Places", "Incorporated Places Labels"),
    options = WMSTileOptions(format = "image/png", transparent = TRUE)
  )

```



